---
title: "A Survival Model in Stan"
author: "Eren Metin Elçi"
date: 2018-09-03
output:
  tufte::tufte_html: default
  tufte::tufte_handout: default
---

The goal of this short case study is to demonstrate the essentials of a Bayesian workflow using the probabilistic programming language Stan. The dataset we consider here is known as the mastectomy dataset^[This case study is motivated by Austin Rochford's related PyMC3 [blog post](https://austinrochford.com/posts/2015-10-05-bayes-survival.html). See also [this](https://docs.pymc.io/notebooks/bayes_param_survival_pymc3.html) updated version, now part of the official PyMC3 documentation. The same dataset was also studied by the same author in [this](https://docs.pymc.io/notebooks/survival_analysis.html) related case study.]. As the title already suggest, we will implement the simplest, and probably most commonly used survival model, also sometimes known as (Cox's) proportional hazard model. 

The beauty or advantage of the Bayesian framework is that we can avoid any technicality or approximation due to what is known as *ties* in the dataset, usually encountered in real datasets and which need special treatments/considerations in frequentist frameworks (see e.g. `coxph`). This is because we model the baseline hazard explicitly and hence do not need to revert to what is known as a pseudo-likelihood^[For a good and concise description of the frequentist approach and the utilized pseudo-likelihood, see chapter 9.4 and the related appendix in [Computer Age Statistical Inference](https://web.stanford.edu/~hastie/CASI/index.html) by Efron and Hastie.].

Now, let's first have a look at the data:

```{r, include=FALSE}
library(HSAUR)
library(tibble)
library(dplyr)
data("mastectomy")
df <- as.tibble(mastectomy)
df <- df %>% mutate(metastized=as.double(metastized=="yes"))
```

```{r, echo=FALSE}
knitr::kable(
  head(df,10), caption = 'A mastectomy dataset'
)
```

More precisely, each row in the dataset represents observations from a woman diagnosed with breast cancer that underwent a mastectomy. 
  
  - The column `time` represents the time (in months) post-surgery that the woman was observed. 
  - The column `event` indicates whether or not the woman died during the observation period. 
  - The column `metastized` represents whether the cancer had metastized^[The cancers are classified as having metastized or not based on a histochemical marker.] prior to surgery.

# Some math (some notation)

Central to survival models is the survival function $S(t)$ defined as 

$$ S(t) = \mathbb{P}[T>t] = e^{-H(t)} $$
Here $T$ is a the survival time of an individual and thus $T>t$ denotes the event that the patient or individual survived beyond time $t$. $H(t)$ is known as the cumulative hazard and can be shown to be given by

$$
H(t) = \int_{0}^{t}{\rm d}u~\lambda(u)
$$

Here we introduced^[We discard the dependence on latent parameters such as $\mathbf{\beta}$ or $\gamma$ below, for the sake of readability.] the hazard rate $\lambda(t)$

$$
h(t;\mathbf{x}) = h_0(t) e^{\mathbf{x}'\cdot\boldsymbol{\beta}}
$$

Here $\mathbf{x}$ is a vector of covariates describing an individual^[In our mastectomy dataset it is simply a scalar indicator corresponding to the column `metastized` above.]. The above makes it apparent why such models are often referred to as *proportional* hazard models. Further, we make the assumption that the baseline hazard $h_0$ fulfills 

$$h_0(t) = h_0.$$

Our Bayesian analysis therefore has the unknown parameters $\boldsymbol{\beta}$ and $h_0$ where we parametrize the latter as $h_0 = e^\gamma$. 
Note that the above implies (or is equivalent to) $T$ having an exponential law with rate parameter equal to $\exp{(\mathbf{x}'\cdot \boldsymbol{\beta}+ \gamma)}$.

For the keen reader, try to verify (or convince yourself) that one has in the limit $dt\rightarrow 0$

$$
h(t)dt \doteq \mathbb{P}\left[T\in (t,t+dt) \vert T\geq t\right]
$$


# Stan

Now let's get our hands dirty (or actually our keyboard) and start specifying our corresponding generative model in Stan!

## Data block

Here we define precisely the type and dimensions of data provided externally^[E.g. via rstan, pystan or cmdstan.] to Stan.
```{rstan  data_block}
data {
    int<lower=0> N_uncensored;                                      
    int<lower=0> N_censored;                                        
    int<lower=1> NC;                                                
    matrix[N_censored,NC] X_censored;                               
    matrix[N_uncensored,NC] X_uncensored;                           
    vector<lower=0>[N_censored] times_censored;                          
    vector<lower=0>[N_uncensored] times_uncensored;                      
}
```

## Parameters block

Here we define all parameters that we wish to infer.

```{rstan  parameters_block}
parameters {
    vector[NC] betas;                                     
    real intercept;                                 
}
```

Note that `betas` corresponds to $\boldsymbol{\beta}$ and `intercept` to $\gamma$.

## Model block

Here we define the likelihood and priors. Before we do so, I'd like to quote [Jonah Gabry](http://iserp.columbia.edu/people/jonah-gabry):

> "Choosing priors is about including information while allowing the chance of being wrong."

In this sense, let's hack the model block:

```{rstan model_block}
model {
    betas ~ normal(0,2);                                                            
    intercept ~ normal(-5,2);                                                     
    target += exponential_lpdf(times_uncensored | exp(intercept+X_uncensored*betas)); 
    target += exponential_lccdf(times_censored | exp(intercept+X_censored*betas));  
}
```

To get an intuition for the prior choice of `intercept` or actually $\gamma$, observe that $e^{-\gamma}$ is equal to the mean of baseline exponential (which in the data is around $100$, hence $\gamma\approx 4.6$).

Note that implicit here is the assumption that survival times are mutually independent. 

Moreover, above we use *vectorized* statements, which makes the computation more efficient than using, say, a for loop and iterating over all individuals. This is the main reason why we decided to work with the `_censored` and `_uncensored` suffixes and split the data, as opposed to the variant where one keeps the data together and provides an boolean array/vector specifying which patients have a (right-) censored survival time.

## Model compilation

```{r  model_compile, message=FALSE}
library(rstan)
rstan_options(auto_write = TRUE)
sm <- stan_model("./exponential_survival_simple_ppc.stan")
```

# Data preparation

```{r data_prep, message=FALSE}
N <- nrow(df)
X <- as.matrix(pull(df, metastized))
is_censored <- pull(df,event)==0
times <- pull(df,time)
msk_censored <- is_censored == 1
N_censored <- sum(msk_censored)
```

Combine (couple) all the data into one named list with reference names corresponding precisely to the actual names as defined in the data block of our Stan model.
```{r stan_list, message=FALSE}
stan_data <- list(N_uncensored=N-N_censored, 
                  N_censored=N_censored, 
                  X_censored=as.matrix(X[msk_censored,]),
                  X_uncensored=as.matrix(X[!msk_censored,]),
                  times_censored=times[msk_censored],
                  times_uncensored = times[!msk_censored],
                  NC=ncol(X)
)
```

# Fitting the model
```{r stan_fit, message=FALSE}
fit <- sampling(sm, data=stan_data, seed=42, chains=4, cores=2, iter=4000)
```

# Inspecting results

Consider especially the `ess` and `rhat` columns below, which correspond to the effective sample size and the potential scale reduction statistics. In a nutshell, `rhat` should be very close to $1$ which indicates that the chain(s) mixed (converges) and `ess` should be as close as possible to the total number of MCMC iterations^[Starting in Stan 2.18 `ess` can in fact be larger than the number of MCMC iterations, essentially due to what is known as anti-correlations (yes NUTS and HMC can sometimes be unbelievable super-efficient!). For more details on the two quantities see the section [General MCMC diagnostics](http://mc-stan.org/bayesplot/articles/visual-mcmc-diagnostics.html#general-mcmc-diagnostics) in the bayesplot vignette [Visual MCMC diagnostics using the bayesplot package
](http://mc-stan.org/bayesplot/articles/visual-mcmc-diagnostics.html). For a detailed example regarding the updates on `ess` and `rhat` in 2.18 I can highly recommend Aki Vehtari's [Rank-normalized split-Rhat and relative efficiency estimates](https://avehtari.github.io/rhat_neff/rhat_neff.html)] (excluding burn-in).

```{r inspect_fit_mcmc, echo=FALSE}
library(broom)
knitr::kable(
 tidy(fit, pars=c("intercept", "betas[1]"), conf.int = TRUE, estimate.method = "median",rhat=TRUE, ess=TRUE),
 caption = 'Posterior summary'
)
```

```{r inspect_fit_nuts, echo=FALSE}
check_divergences(fit)
check_treedepth(fit)
```

## Visual inspection of the posterior

```{r prep_visual, message=FALSE}
library(bayesplot)
library(survival)
post <- as.array(fit)
fit_cox <- coxph( Surv(time, event)~metastized, data=df)
coef_cox <- coef(fit_cox)
se_cox <- sqrt(fit_cox$var)
```

### Kernel density plots of posterior draws with chains separated but overlaid on a single plot.

```{r visual_inspect_1, message=FALSE}
mcmc_dens_overlay(post, pars=c("betas[1]", "intercept")) 
```

### Plots of uncertainty intervals computed from posterior draws with all chains merged.

The three vertical lines below (from left to right) correspond to the frequentist's^[Using the `coxph` routine in the survival package, see the code of this RMarkdown.] point estimate minus the standard error, the point estimate and the point estimate plus the standard error of the regression coefficient, respectively. 

```{r visual_inspect_2, message=FALSE}
mcmc_intervals(post, pars=c("betas[1]", "intercept")) + 
  vline_at(c(coef_cox-se_cox, coef_cox, coef_cox+se_cox))
```

### Pairs plot
```{r visual_inspect_3, message=FALSE}
color_scheme_set("red")
mcmc_pairs(post, pars=c("betas[1]", "intercept"))
```

### Hex plot
```{r visual_inspect_4, message=FALSE}
color_scheme_set("gray")
mcmc_hex(post, pars=c("betas[1]", "intercept"))
```

### Trace plot
```{r visual_inspect_5, message=FALSE}
color_scheme_set("mix-blue-red")
mcmc_trace(post, pars=c("betas[1]", "intercept"),
           facet_args = list(ncol = 1, strip.position = "left")
           )
```

# Kaplan Meier plot
```{r kaplan_meier, echo=FALSE}
library(purrr)
lambda_base = as.vector(exp(post[,,"intercept"]))
lambda_metastized = lambda_base * as.vector(exp(post[,,"betas[1]"]))
times_plot <- seq(0, max(times), length.out = 1000)
surv_base <- map(times_plot, ~exp(-.*lambda_base))
surv_base_qtl <- map(surv_base, ~quantile(., probs=c(0.05, .975)))
surv_metastized <- map(times_plot, ~exp(-.*lambda_metastized))
surv_metastized_qtl <- map(surv_metastized, ~quantile(., probs=c(0.05, .975)))

ggplot(
  data=bind_rows(
  tibble(
    t = times_plot,
    metastized=FALSE,
    surv=map_dbl(surv_base, median),
    surv_low=map_dbl(surv_base_qtl, ~.[["5%"]]),
    surv_up=map_dbl(surv_base_qtl, ~.[["97.5%"]])
  ),
  tibble(
    t = times_plot,
    metastized=TRUE,
    surv=map_dbl(surv_metastized, median),
    surv_low=map_dbl(surv_metastized_qtl, ~.[["5%"]]),
    surv_up=map_dbl(surv_metastized_qtl, ~.[["97.5%"]])
  )),aes(t))+
  geom_ribbon(aes(ymin = surv_low, ymax = surv_up, fill=metastized), alpha=.3) +
  geom_line(aes(y = surv, color=metastized))
```

# Posterior predictive checks

Posterior predictive checks^[See the excellent bayesplot vignette [Graphical posterior predictive checks using the bayesplot package](https://cran.r-project.org/web/packages/bayesplot/vignettes/graphical-ppcs.html) for an introdution and practical instructions.] constitute a family of powerful methods to scrutinize relevant aspects of your model.

Below we run various posteriod predictive checks constrained to the instances (individuals) that came with non-censored survival times:

```{r ppc, message=FALSE}
surv_times_rep <- as.matrix(post[,1,sprintf("times_censored_sampled[%d]", 1:stan_data$N_uncensored)]) 
surv_times_train <- times[!msk_censored]
###########################################################################################
color_scheme_set("brightblue")
ppc_dens_overlay(surv_times_train, surv_times_rep[1:500,])
ppc_stat(surv_times_train, surv_times_rep, stat = "mean")
ppc_stat(surv_times_train, surv_times_rep, stat = "sd")
ppc_stat(surv_times_train, surv_times_rep, stat = "max")
ppc_stat(surv_times_train, surv_times_rep, stat = "min")
```

As it turns out our model essentially suffers from overdispersion and a tendency for too large survival times. Potential ways to improve this, would be to consider accelerated failure time models, semiparametric base-line hazards or more general parametric survival models, like the Royston & Parmar^[[Flexible parametric proportional-hazards and proportional-odds models for censored survival data, with application to prognostic modelling and estimation of treatment effects.](https://www.ncbi.nlm.nih.gov/pubmed/12210632). A related case study will follow soon.] based family of models.

## Generated Quantities Block

The careful reader probably noticed that we did not specify how to obtain the parameter vector `times_censored_sampled` required for the posterior predictive checks. This can be achieved by appending the `generated quantities` block below to our Stan model:

```{rstan generated_quantities}
generated quantities {
    vector[N_uncensored] times_censored_sampled;
    for(i in 1:N_uncensored) {
        times_censored_sampled[i] = exponential_rng(exp(intercept+X_uncensored[i,]*betas));
    }
}
```